---
title: "Teaching Math Effectively: Insights from LSAY and PISA on Primary Teacher Proficiency"
author: Tashya Sathyajit (29672732)
format: pdf
number-sections: true
execute: 
  echo: false
  warning: false
  message: false
  wordcount: true
toc: true
---

```{r}
library(dplyr)
library(kableExtra)
```

# Abstract

This report investigates the mathematics performance of individuals who became primary school teachers in Australia, using the Longitudinal Surveys of Australian Youth (LSAY) and the Program For International Student Assessment (PISA) datasets. Results show that teachers consistently have a higher average math proficiency than non-teachers over time, with better performance across PISA testing years. However, they tend to score lower than professionals in fields like medicine, law, and engineering. Limitations include disproportionate teacher counts in some PISA years, potentional attrition bias, confounding variables such as socioeconomic status, and the evolving nature of math proficiency over time.

# Background and motivation

Australian school students are consistently under performing in mathematics, in particular almost half of Australia's 15 year-old students are failing to achieve national standards in mathematics with the nation now more than four years behind the world's top-performing [jurisdiction in maths](https://www.theage.com.au/national/victoria/australian-school-students-more-than-four-years-behind-in-maths-20231204-p5eosl.html).



# Objectives and significance

The primary objectives of this analysis was to utilise longitudinal survey data to understand  the mathematics proficiency of students who eventually became primary school teachers in Australia. Using the LSAY and PISA datasets, the analysis seeks to address four main questions:

1. **Teacher performance over time:** How well did individuals who became primary school teachers perform in mathematics across the years?
2. **Comparison to other Professionals:** How are they performing in comparison to other 'professionals'?
3. **Comparison to the General Population:** How do teachers' math scores compare to the rest of the sample (non primary school teachers)?
4. **Proficiency standards:** How many teachers meet or exceed the mathematics proficiency standard in each year and over time?

This analysis is significant as it provides critical insights into the math skills of individuals who become primary school teachers, a factor that can directly impact student outcomes. By identifying trends and proficiency gaps, the analysis can allow for a deeper understanding of the performance of primary school teachers in Australia and inform potential

# Methodology

The aim of the project was to understand the performance of primary school teachers in Australia in mathematics, utilising the Longitudinal Surveys of Australian Youth [(LSAY)](https://www.lsay.edu.au/) and Programme for International Student Assessment [(PISA)](https://www.oecd.org/en/about/programmes/pisa.html) datasets. `Intsvy` and `rrepest` were used to determine accurate summary statistics.

We wanted to 


### The Data

The analysis required data which retained the demographic information from LSAY with the corresponding PISA scores for each observation. To achieve this, LSAY data post 2003 was used as participants were recruited from schools that also took part in the PISA. These years included: 2003, 2006, 2009 and 2015.

In the dataset each row represented a student, and each column had demographical data encoded into variable names, PISA scores and weights.

A dummy data frame was created for better understanding:

```{r}
# creating a dummy data frame as data is private
dummy_data <- data.frame(
  STIDSTD = sample(1:5, 10, replace = TRUE),
  SECTOR = sample(1:2, 10, replace = TRUE),
  SEX = sample(1:2, 10, replace = TRUE),
  PV1MATH = runif(10, 400, 600),
  PV2MATH = runif(10, 400, 600),
  ST38Q03 = sample(1:5, 10, replace = TRUE),
  ST38Q04 = sample(1:5, 10, replace = TRUE),
  ST38Q05 = sample(1:5, 10, replace = TRUE),
  w_fstr1 = runif(10, 0.5, 1.5)
)

# Display the dummy dataframe with adjusted styling to fit the page
dummy_data %>%
  kable("latex", booktabs = TRUE, longtable = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  column_spec(1:5, width = "5em") %>%
  row_spec(0, bold = TRUE, background = "#D3D3D3")
```


# Results

# Discussion

# Conclusion

